Partitioned Local Depth (PaLD) Clustering Analyses in R

### 2022-03-22 r-journal/manuscript.Rmd
update paper

* Describe PaLD (Ken & Kate?) [@berenhaut2022social]

We present a new package, \CRANpkg{pald}, for calculating partitioned local depth (PaLD) probabilities, implementing clustering analyses, and creating data visualizations to represent the clusters. This paper will describe how to use the package as well as walk through two examples.

>>>>

Partitioned Local Depths (PaLD) is a framework for a holistic consideration of the community structure of distance-based data.  Leveraging a socially inspired perspective, the method provides network-based community information which is founded on a new measure of local depth and pairwise cohesion (partitioned local depth).  The method does not require distributional assumptions, optimization criteria, nor extraneous inputs.  A complete description of the perspective, together with a discussion of the underlying social motivation, theoretical results, and applications to additional data sets is provided in \citet{berenhaut2022social}.

Building on existing approaches to (global) depth, local depth expresses features of centrality as an interpretable probability which is free of parameters and robust to outliers.  Then, partitioning the probability which defines local depth, we obtain a measure of cohesion between pairs of points.  Both local depth and cohesion reflect aspects of relative position (rather than absolute distance) and provide a straightforward way to account for varying density across the space.  Specifically, provided that two sets are separated (in the sense that the minimum between-set distance is greater than the maximum within-set distance), cohesion is invariant under the contraction and dilation of the distances within each set.  This property may be particularly valuable when one has reason to believe that the magnitude of distances varies across the space.

As cohesion captures a sense of the relationship strength between points, we can then visualize the resulting community structure with a network whose edges are weighted by (mutual) cohesion. The underlying social framework motivates an elegant threshold for distinguishing between strongly and weakly cohesive pairs.

As seen throughout this paper, we display the network obtained from cohesion using a force-directed graph drawing algorithm and emphasize the strong ties (colored by connected component).  We refer to the connected components of the network of strong ties as "clusters."  Note that to qualify as a cluster in this definition, one may not have any strong ties with those outside the cluster, and thus the existence of disjoint groups is a strong signal for separation.  Here, clusters are identified without additional user inputs nor optimization criteria.  If one wishes to further break the community graph into groups, one may use community detection methods for networks (such as spectral clustering or the Louvain algorithm).  Though only briefly considered here, one may also use the collection of strong ties in place of (weighted) k-nearest neighbors in settings such as classification and smoothing.  Overall, the structural information obtained from local depth, cohesion and community graphs can provide a holistic perspective on the data which does not require the use of distributional assumptions, optimization criteria nor additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned Local Depths (PaLD) probabilities, implementing clustering analyses, and creating data visualizations to represent the clusters. This paper will describe how to use the package as well as walk through two examples.

### 2022-03-22 r-journal/manuscript.tex
update paper

\begin{itemize}
\tightlist
\item
  Describe PaLD (Ken \& Kate?) \citep{berenhaut2022social}
\end{itemize}

We present a new package, \CRANpkg{pald}, for calculating partitioned
local depth (PaLD) probabilities, implementing clustering analyses, and
creating data visualizations to represent the clusters. This paper will
describe how to use the package as well as walk through two examples.

@@

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH dcdcffa UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from dcdcffa (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 64a847b:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}

>>>>

Partitioned Local Depths (PaLD) is a framework for a holistic
consideration of the community structure of distance-based data.
Leveraging a socially inspired perspective, the method provides
network-based community information which is founded on a new measure of
local depth and pairwise cohesion (partitioned local depth). The method
does not require distributional assumptions, optimization criteria, nor
extraneous inputs. A complete description of the perspective, together
with a discussion of the underlying social motivation, theoretical
results, and applications to additional data sets is provided in
\citet{berenhaut2022social}.

Building on existing approaches to (global) depth, local depth expresses
features of centrality as an interpretable probability which is free of
parameters and robust to outliers. Then, partitioning the probability
which defines local depth, we obtain a measure of cohesion between pairs
of points. Both local depth and cohesion reflect aspects of relative
position (rather than absolute distance) and provide a straightforward
way to account for varying density across the space. Specifically,
provided that two sets are separated (in the sense that the minimum
between-set distance is greater than the maximum within-set distance),
cohesion is invariant under the contraction and dilation of the
distances within each set. This property may be particularly valuable
when one has reason to believe that the magnitude of distances varies
across the space.

As cohesion captures a sense of the relationship strength between
points, we can then visualize the resulting community structure with a
network whose edges are weighted by (mutual) cohesion. The underlying
social framework motivates an elegant threshold for distinguishing
between strongly and weakly cohesive pairs.

As seen throughout this paper, we display the network obtained from
cohesion using a force-directed graph drawing algorithm and emphasize
the strong ties (colored by connected component). We refer to the
connected components of the network of strong ties as ``clusters.'' Note
that to qualify as a cluster in this definition, one may not have any
strong ties with those outside the cluster, and thus the existence of
disjoint groups is a strong signal for separation. Here, clusters are
identified without additional user inputs nor optimization criteria. If
one wishes to further break the community graph into groups, one may use
community detection methods for networks (such as spectral clustering or
the Louvain algorithm). Though only briefly considered here, one may
also use the collection of strong ties in place of (weighted) k-nearest
neighbors in settings such as classification and smoothing. Overall, the
structural information obtained from local depth, cohesion and community
graphs can provide a holistic perspective on the data which does not
require the use of distributional assumptions, optimization criteria nor
additional user inputs.

We present a new package, \CRANpkg{pald}, for calculating Partitioned
Local Depths (PaLD) probabilities, implementing clustering analyses, and
creating data visualizations to represent the clusters. This paper will
describe how to use the package as well as walk through two examples.

@@

\begin{Schunk}
\begin{Sinput}
graphs <- community_graphs(cohesion)
graphs[["G_strong"]]
\end{Sinput}
\begin{Soutput}
#> IGRAPH b3faedc UNW- 6 3 -- 
#> + attr: name (v/c), weight (e/n)
#> + edges from b3faedc (vertex names):
#> [1] A--B A--E D--F
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
french_neighbors <- igraph::neighbors(cognate_graph_strong, "French")
french_neighbors
\end{Sinput}
\begin{Soutput}
#> + 8/87 vertices, named, from 46d19b2:
#> [1] Italian         Ladin           Provencal       Walloon        
#> [5] French_Creole_C French_Creole_D Spanish         Catalan
\end{Soutput}
\end{Schunk}
